---
title: "Intro to SQL in R"
author: "Ivan Leung"
date: "8/4/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: paper
    css: css/hideOutput.css
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(kableExtra)
library(knitr)
library(rmarkdown)
library(RSQLite)
library(DBI)
library(dplyr)
library(dbplyr)

PrettyTable <- function(data) {
  data %>% 
    kable() %>% 
    kable_styling(bootstrap_options = c("striped", "hover",
                                        "condensed", "responsive"),
                  full_width = FALSE)
}

#Color Format
ColFmt <- function(x, color = 'red') {
  outputFormat <- knitr::opts_knit$get("rmarkdown.pandoc.to")
  if (outputFormat == "latex") {
    paste("\\textcolor{", color, "}{", x, "}", sep = "")
  } else if (outputFormat == "html") {
    paste("<font color='", color, "'>", x, "</font>", sep = "")
  } else {
    x
  }
}

```

# Create a Database

First create a database and connect to it.

For this guide, a `SQLite` database will be used.

```{r, eval=FALSE}
# if the following packages are not installed yet, please install them with
# install.packages('package_name')

library(RSQLite)
library(DBI)
library(dplyr)
library(dbplyr)
```

Database may be created as a temporary or permanent entity, depending on use case. For this guide, a temporary one will be used.

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
```

For a permanent database, substitute the `dbname` argument with the desired path and name of the to-be-created-database.

In the code below, a database called `my-db.sqlite` will be created in current working directory.

```{r, eval=FALSE}
mydb <- dbConnect(RSQLite::SQLite(), "my-db.sqlite")
dbDisconnect(mydb)
```

In the real world, a dedicated database admin team will have created the database for users to connect to, with built-in user authentication; an example is shown below:

```{r, eval=FALSE}
real_con <- DBI::dbConnect(RSQLite::SQLite(),
                      host = "12.34.56.78",
                      user = rstudioapi::askForPassword("Enter username"),
                      password = rstudioapi::askForPassword("Enter password")
)
```

# SQL in R

A database connection `con` was created above; let's load in data into it.

```{r}
dbWriteTable(con, "mtcars", mtcars)
dbWriteTable(con, "iris", iris)
```

Check to confirm data is in database.

```{r}
dbListTables(con)
```

## Native SQL

If one is familiar with SQL already, direct querying may be done.

```{r}
dbGetQuery(con, 'SELECT * FROM iris WHERE "Sepal.Length" < 4.6')

dbGetQuery(con, 'SELECT distinct gear, carb FROM mtcars')
```

## `dplyr` SQL

If one is not familiar with SQL (though it's highly recommend to learn `r emo::ji('smile')`), then reply on `dplyr` (with some catch...details below).

The codes below are taken from the [dbplyr vignette](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html){target="_blank"}.

The `copy_to()` function from `dplyr` may be used to load data into the database as well.

Below, the `flights` data is loaded in.

```{r}
copy_to(con, 
        df = nycflights13::flights, 
        name = "flights",
        # create table indices for faster query
        indexes = list(
          c("year", "month", "day"), 
          "carrier", 
          "tailnum",
          "dest"
        )
)
```

Confirm `flights` data is in database.

```{r}
dbListTables(con)
```

### Connecting to Data Table in Database

Here, we connect to the data table in the database, using `tbl()` from `dplyr`.

```{r}
flights_db <- tbl(con, "flights")
```

Note that this is a _lazy_ connection! More on this below.

Notice the `[?? x 19]` in the output header for the code below.

```{r}
flights_db
```

### SQL query via `dplyr`

Examples below illustrate the convenience of using `dplyr` syntax for SQL query.

#### Example 1

```{r}
flights_db %>%
  select(year:day, dep_delay, arr_delay)
```

Above is the same as running:

```{r, eval=FALSE}
dbGetQuery(con, "
SELECT `year`, `month`, `day`, `dep_delay`, `arr_delay`
FROM flights")
```

#### Example 2

```{r}
flights_db %>% filter(dep_delay > 240)
```

Above is the same as running:

```{r, eval=FALSE}
dbGetQuery(con, "
SELECT *
FROM flights
WHERE (`dep_delay` > 240.0)")
```

#### Example 3

```{r}
flights_db %>%
  group_by(dest) %>%
  summarise(delay = mean(dep_time))
```

Above is the same as running:

```{r, eval=FALSE}
dbGetQuery(con, "
SELECT `dest`, AVG(`dep_time`) AS `delay`
FROM flights
GROUP BY `dest`")
```

### Lazy Query
As per `dbplyr` vignette, queries made via `dplyr` tries to be as lazy as possible:

- It never pulls data into R unless you explicitly ask for it.

- It delays doing any work until the last possible moment: it collects together everything you want to do and then sends it to the database in one step.

This is demonstrated below.

Again, notice the `[?? x 3]` in the output header for the code below.

```{r}
tailnum_delay_db <- flights_db %>% 
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay),
    n = n()
  ) %>% 
  arrange(desc(delay)) %>%
  filter(n > 100)

tailnum_delay_db
```

Lazy query has its disadvantages.

Because there’s generally no way to determine how many rows a query will return unless you actually run it, `nrow()` is always `NA`.

```{r, eval=FALSE}
nrow(tailnum_delay_db)
#> [1] NA
```

Consequently, because you can’t find the last few rows without executing the whole query, you can’t use tail()

```{r, eval=FALSE}
tail(tailnum_delay_db)
#> Error: tail() is not supported by sql sources
```

But other database metadata that are _fixed_ may be obtained.

```{r}
colnames(tailnum_delay_db)
```

### Behind the (SQL) Scene

`dplyr` kindly translated our R code to SQL equivalent, which may be inspected with `show_query()`.

```{r}
tailnum_delay_db %>% show_query()
```

Finally, data is pulled into memory via `collect()`.

Notice that the output header now shows actual data dimensions - `1,201 x 3`.

```{r}
tailnum_delay <- tailnum_delay_db %>% collect()
tailnum_delay
```

### Table Joins

Joining is supported too, with a catch.

In the demonstration below, the `flights` data from before is to be joined with `airlines` data via the `carrier` code column.

First load the data into database.

```{r}
copy_to(con, nycflights13::airlines, "airlines")
```

Check to confirm.

```{r}
dbListTables(con)
```

Create a table connection.

```{r}
airlines_db <- tbl(con, "airlines")
airlines_db
```

Finally, the join.

```{r}
flights_db %>% 
  distinct(carrier) %>% 
  left_join(airlines_db, by = "carrier")
```

Yay, it works as expected!

Now the twist...

In the real world, users sometimes have to blend data from different databases. To facilitate this flow of data, _database links_ are often created by database admins.

Let's see an example below where the `airlines` data is now located in a different database.

We'll repeat the above steps, *but* on a new temporary database, called `con2`.

```{r}
con2 <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con2, nycflights13::airlines, "airlines")

# check to confirm
dbListTables(con2)

# create table connection
airlines_db2 <- tbl(con2, "airlines")
airlines_db2
```

Now join the data tables as before.

```{r, eval=FALSE}
flights_db %>% 
  distinct(carrier) %>% 
  left_join(airlines_db2)

#> Joining, by = "carrier"
#> Error: `x` and `y` must share the same src, set `copy` = TRUE (may be slow)
```

It fails, for the reason noted above - these data tables sit on different databases.

As a workaround, one may use the `copy` argument and set it to `TRUE`.

```{r}
flights_db %>% 
  distinct(carrier) %>% 
  left_join(airlines_db2, copy = TRUE)
```

Be warn that this workaround is not ideal for large data tables. In such cases, a solution implemented within the database itself (e.g., stored procedure) is recommended.
  
## Debugging in `dplyr` SQL

There is ongoing work to translate `dplyr` syntax into SQL equivalent, but there are just cases where it's impossible.

Below are some examples on how to check and improvise.

First, create a fake database dataframe.

```{r}
mf <- memdb_frame(x = letters, y = seq(letters))
```

Any function that `dbplyr` doesn’t know about will be left as is

```{r}
mf %>% 
  mutate(z = foofify(x, y)) %>% 
  show_query()
```

Other functions are translated as a _hybrid_.

In the example below, one would have used `stringr::str_detect()`, for example; however, different SQL flavours (e.g., Microsoft SQL, SQLite) handles regular expressions differently, or not handle it at all.

As a workaround, such hybrid achieves the best of both worlds.

```{r}
mf %>% 
  filter(x %LIKE% "%foo%") %>% 
  show_query()
```

A R `paste()` equivalent in SQL (i.e. `||`).

```{r}
mf %>% 
  transmute(z = x %||% y) %>% 
  show_query()
```

Another flexibility of `dbplyr` is it allows mixing of `dplyr` and native SQL using `sql()`.

Note that in code below, `filter()` is translated to `WHERE` in SQL, while `CAST()` stays as-is.

```{r}
mf %>% 
  filter(y > 10) %>% 
  transmute(factorial = sql("CAST(x AS FLOAT)")) %>% 
  show_query()
```

Check out [translated functions]( https://cran.r-project.org/web/packages/dbplyr/vignettes/translation-function.html){target="_blank"} for a list of functions that `dbplyr` currently supports.



